Gender, Education, and Mental Depression

### 2024-10-21 paper/paper.qmd
Adds Chritina's initial text suggestions and removes some unnecessary filler from the starter_folder.

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(boot)
library(broom.mixed)
library(collapse)
library(dataverse)
library(gutenbergr)
library(janitor)
library(knitr)
library(marginaleffects)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(tidyverse)
df <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))

```

```{r}
#| warning: false
#| message: false
labeled_df |>
  ggplot(aes(x = label_education, fill = label_outcome)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(label_sex)) +
  theme_minimal() +
  labs(
    x = "Highest education",
    y = "Number of respondents",
    fill = "Mental Depression"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "bottom")
```
```{r}
#| warning: false
#| message: false
mental_illness <- readRDS(file = here::here("models/mental_illness.rds"))

modelsummary(
  list(
    "Mental Illness Exist" = mental_illness
  ),
  statistic = "mad"
  )
```

# Introduction

@@

@@

Overview text

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

@@

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

@@

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

@@

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

@@

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

@@

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

@@

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

>>>>

---
title: "Gender, Education, and Mental Depression"
subtitle: "Analyzing Census Data Through Statistical Methods"
author: 
  - Nibras Ar Rakib
  - Christina Nguyen
  - Marcus Emmanuel Barnes
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(boot)
library(broom.mixed)
library(collapse)
library(dataverse)
library(gutenbergr)
library(janitor)
library(knitr)
library(marginaleffects)
library(modelsummary)
library(rstanarm)
library(tidybayes)
library(tidyverse)
# df <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
labeled_df <- read.csv(here::here("data/02-analysis_data/model_data_sample_2k.csv"))
```
# Introduction
Key questions include: Does higher education levels correlate with lower rates of depression? Are women disproportionately affected by depression compared to men at the same educational level?

@@

The data for this study is drawn from national census data, which includes variables for gender (male/female), education level (from grade 1 to doctoral degree), and mental depression (no/yes). The sample consists of individuals age (FILL THIS IN AND WHY WE CHOSE THIS).

```{r}
#| label: fig-education
#| fig-cap: Education
#| warning: false
#| message: false
#| echo: false
labeled_df |>
  ggplot(aes(x = education, fill = outcome)) +
  stat_count(position = "dodge") +
  facet_wrap(facets = vars(sex)) +
  theme_minimal() +
  labs(
    x = "Highest education",
    y = "Number of respondents",
    fill = "Mental Depression"
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "bottom")
```
The data for this study is drawn from national census data, which includes variables for gender (male/female), education level (from grade 1 to doctoral degree), and mental depression (no/yes). The sample consists of individuals age (FILL THIS IN AND WHY WE CHOSE THIS).

@@

### Gender and Mental Health
Research consistently shows that women report higher rates of depression than men (CITE). Biological, social, and psychological factors contribute to this disparity. Women may experience unique stressors, such as role strain from balancing work and family obligations, as well as being subject to gender-based discrimination (CITE). These pressures can exacerbate mental health issues, leading to higher rates of depression among women, particularly those in lower socioeconomic or educational strata.

### Educational Attainment and Mental Health
Education is often seen as a protective factor against mental health issues. Studies suggest that higher levels of education correlate with improved mental health outcomes (CITE). Educated individuals tend to have better access to mental health resources, higher socioeconomic status, and greater job security, all of which contribute to lower rates of depression (CITE). (DO WE WANT DATA ON HEALTH CARE COVERAGE/INSURANCE HERE?) Conversely, those with lower education levels may face more financial strain, reduced employment opportunities, and limited access to healthcare, increasing their vulnerability to depression.

### Intersection of Gender, Education, and Depression
Despite the protective nature of education, the gender gap in mental health persists even at higher levels of educational attainment. Some studies suggest that women with high education levels still experience depression due to external stressors, including workplace discrimination and societal expectations (CITE). Additionally, the mental health benefits of education may differ between genders, with men benefiting more from the social status and economic opportunities that education provides, while women may continue to face challenges related to gender roles (cite).

## Methodology

### Statistical Techniques
This study employs the following statistical techniques:

* Descriptive statistics : These will summarize the distribution of education levels and depression rates across genders.
* Logistic regression: We will use logistic regression to examine how education level and gender predict the likelihood of experiencing depression, controlling for other demographic factors like age and income.

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset (that is talk about the data was collected).

@@

@@

@@

@@

@@

We expect a positive relationship between education level and

@@

@@

## Gender Differences in Depression {#discussion-first-point}
The finding that women report higher rates of depression across all educational levels aligns with existing literature. This disparity can be attributed to a variety of factors, including biological differences in stress response and the disproportionate burden women face in managing work-life balance (CITE). Additionally, women may be more likely to seek help and report depressive symptoms, contributing to higher rates in survey data (CITE).

## Education as a Protective Factor {#discussion-second-point}
The results support the hypothesis that education serves as a protective factor against depression. Those with higher education levels enjoy greater access to economic and social resources, reducing the risk of depression (CITE). However, the persistent gender gap in mental health, even among those with higher education, suggests that education alone is insufficient to fully mitigate the effects of gender-based stressors and societal pressures on women (CITE AND CHECK MY FACTS HERE).

## Intersectionality and Mental Health
The intersection of gender and education highlights the complexity of mental health outcomes. Women with lower education levels are doubly disadvantaged, experiencing both the socioeconomic limitations of low education and the additional burdens associated with gender. Future research should explore how other intersecting factors, such as race and socioeconomic status, further influence the relationship between education and mental health.

@@

# Conclusion
This paper has demonstrated that both education and gender play crucial roles in shaping mental health outcomes, particularly depression.[ETC FILL THIS IN]

@@

@@
